---
title: "Code for the Analysis of the Clinical Trial on D-penicillamine efficiency to treat Primary Biliary Cirrhosis performed by the Mayo Clinic in the 70’s"
output: pdf_document
---

## Description of the Mayo Clinic Primary Biliary Cirrhosis Data

id:	case number
time: 	number of days between registration and the earlier of death, transplantion, or study analysis in July, 1986
    => Transformed in months for a better comprehension
    
status:	status at endpoint, 0/1 for censored, dead
    => transplanted patients are pooled wither censored one
        
trt:	1/2/NA for D-penicillmain, placebo, not randomised
    => Rename in 'Drug', 'Placebo', 'Not_rand'

sex:	m/f
age:	in years
stage:	histologic stage of disease (needs biopsy)

ascites:	presence of ascites
edema:	0 no edema, 
        0.5 untreated or successfully treated
	      1 edema despite diuretic therapy
hepato:	presence of hepatomegaly or enlarged liver
spiders:	blood vessel malformations in the skin

albumin:	serum albumin (g/dl)
alk.phos:	alkaline phosphotase (U/liter)
ast:	aspartate aminotransferase, once called SGOT (U/ml)
bili:	serum bilirunbin (mg/dl)
chol:	serum cholesterol (mg/dl)
copper:	urine copper (ug/day)
trig:	triglycerides (mg/dl)

platelet:	platelet count
protime:	standardised blood clotting time



## library import

```{r}
library(dplyr)
library(survival)
```


## Data importation and transformation

```{r}
data <- pbc
# When transplanted -> Censored
# Survive is now 1
data <- mutate(data, status = ifelse(status == 1, 0,
                              ifelse(status == 2, 1, status)))

# Replace treatment value 0/1 by drug/placebo for better comprehension later
data <- mutate(data, trt = ifelse(trt == 1, 'drug', 
                           ifelse(trt == 2, 'placebo', trt)))
data <- mutate(data, trt = ifelse(is.na(data$trt), 'Not_rand', trt))

# Modification of the 'time' from day to month
data <- mutate(data, time = time/(365.25/12))

data$trt <- as.factor(data$trt)
data$ascites <- as.factor(data$ascites)
data$hepato <- as.factor(data$hepato)
data$spiders <- as.factor(data$spiders)
data$edema <- as.factor(data$edema)
data$stage <- as.factor(data$stage)
#data$status <- as.factor(data$status)

head(data)
summary(subset(data, trt != 'Not_rand'))

```


## Test of the repartition of the variable 'sex' between placebo and D-penicillamin group

```{r}
tmp <- with(subset(data, trt != 'Not_rand' ), table(trt, sex))
tmp

fisher.test(tmp)
```

# Test of the age repartition between placebo and D-penicillamin group

```{r}
# Placebo Patients are younger

placebo <- subset(data, trt == 'placebo', age)
drug <- subset(data, trt == 'drug', age)

boxplot(c(placebo, drug))
t.test(placebo, drug)

# Cox regression model show an effect of the age
cox <- coxph(Surv(time, status) ~ age , data = subset(data, trt != 'Not_rand' ))



```

## Age stratification

```{r}
summary(subset(data, trt != 'Not_rand'))

data <- mutate(data, age_group = ifelse(age < 43, 'young', 
                                 ifelse(age > 56, 'old', 'middle')))

data$age_group <- as.factor(data$age_group)
summary(data)

```

## Cox regression model to test treatment effect considering treatment, age and sex

```{r}
#Cox Regression
cox <- coxph(Surv(time, status) ~ trt + sex + strata(age_group) , data = subset(data, trt != 'Not_rand'))
cox

# Age distribution of the not randomized group
Not_rand <- subset(data, trt == 'Not_rand', age)
summary(Not_rand)
boxplot(c(placebo, drug, Not_rand))

# Kaplan-Meier estimation
surv_fit <- survfit(Surv(time, status) ~ trt , data = data)
plot(surv_fit, col =1:3, xlab='Time in month', ylab='Survival rate')
legend("topright", lty = 1, col = 1:3, legend = names(surv_fit$strata))

# LogRank test
surv_diff <- survdiff(Surv(time, status) ~ trt , data = subset(data, trt != 'Not_rand'))
surv_diff

```



## PART 2

## Creation of a dataset with no missing values to compute the model
```{r}
# Remove rows with missing values
data_model <- data[complete.cases(data),]
```

## Model with only placebo group 
```{r}
data_placebo <- subset(data_model, trt == 'placebo' )

tail(data)
```


## Cox regression model with all the variables on group placebo data
```{r}

model_full_placebo <- coxph(Surv(time, status) ~ sex + age + ascites + hepato + spiders + log(bili) + log(chol) + log(albumin) + log(copper) + log(alk.phos) + log(ast) + log(trig) + platelet + protime, data = data_placebo)

```

## Automatic Selection of the covariates using stepwise method
```{r}
model_placebo <- step(model_full_placebo)
```

## Summary of the 'Best' model

```{r}
model_placebo
```



# As it's not significant, spiders were removed from the placebo model
```{r}
model_placebo <- coxph(formula = Surv(time, status) ~ age + spiders + log(bili), data = data_placebo)

summary(model_placebo)
```


## Martingales residuals for model 1

```{r}

# Calculate Martingales
data_placebo$residual <- residuals(model_placebo, type='martingale')

#Plot martingales
par(mfrow = c(1, 3))
with(data_placebo, {

  plot(age , residual)
  lines(lowess(age, residual), lwd = 2)
  
  plot(bili , residual)
  lines(lowess(bili, residual), lwd = 2)
  
  plot(residual ~ spiders)

})

```

## Deltas-Betas for model 1

```{r}
dfbetas <- residuals(model_placebo, type = 'dfbetas')
data_placebo$dfbetas <- sqrt(rowSums(dfbetas^2))


plot(data_placebo$dfbetas, type = 'h', ylab= 'Deltas-Betas Residuals', xlab= 'Patient ID' )

```


## MODEL 2: Dataset taken into consideration: group Placebo + D-penicillamine

## Cox regression model with all the variables on group placebo data

```{r}
model_full <- coxph(Surv(time, status) ~ sex + age + ascites + hepato + spiders + log(bili) + log(chol) + log(albumin) + log(copper) + log(alk.phos) + log(ast) + log(trig) + platelet + protime, data = data_model)
```

## Automatic Selection of the covariates using stepwise method

```{r}
model <- step(model_full)
```



# Summary of model 2

```{r}
model <- coxph(formula = Surv(time, status) ~ age + log(bili) + log(albumin) + log(log(copper)) + protime, data = data_model)
  
summary(model)
```


## Martingales for model 2

```{r}
# Calculate Martingales
data_model$residual <- residuals(model, type='martingale')

# Plot Martingales for each covariates
par(mfrow = c(2, 3))
with(data_model, {

  plot(age , residual)
  lines(lowess(age, residual), lwd = 2)
  
  plot(bili , residual)
  lines(lowess(bili, residual), lwd = 2)

  plot(albumin , residual)
  lines(lowess(albumin, residual), lwd = 2)
  
  plot(copper , residual)
  lines(lowess(copper, residual), lwd = 2)
  
  plot(protime , residual)
  lines(lowess(copper, residual), lwd = 2)
  

})


```

## Delta-betas for model 2

```{r}
dfbetas <- residuals(model, type = 'dfbetas')
data_model$dfbetas <- sqrt(rowSums(dfbetas^2))


plot(data_model$dfbetas, type = 'h', ylab= 'Delta-Betas Residuals', xlab = 'Patient Id')
abline(h = 0)
```
